---
title: "Introduction to `threemc`"
output: rmarkdown::html_vignette
# toc: true
bibliography: library.bib
csl: plos-one.csl
vignette: >
  %\VignetteIndexEntry{Introduction to `threemc`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette serves as an introduction to `threemc` (or "Matt's model for male 
circumcision"), a Bayesian multilevel spatio-temporal, competing-risks, 
time-to-event model for, unsurprisingly, male circumcision. It models the 
circumcision rate, but also computes circumcision incidence and 
cumulative incidence (i.e. circumcision coverage) [@thomas_multi-level_2021].

Circumcision is split between two types (which form our "competing-risks", 
along with being uncircumcised), medical male circumcision (MMC) and 
traditional male circumcision (TMC). The model is stratified by age, location 
and time, and also includes interaction terms between these variables.  

`threemc` is implemented in Template Model Builder (`TMB`).
TMB is an R package that enables users to flexibly fit latent process/variable 
models to data. It uses automatic differentiation and Laplace approximations to 
estimate posterior distributions for model parameters.
The objective function produced by `TMB` can be called in R, and so can be 
optimised using e.g. `stats::nlminb`. `TMB` is particularly fast in 
comparison to other methods (e.g. INLA) when there are many random effects, 
such as are commonly found in spatio-temporal models like `threemc`. 
[@kristensen_tmb_2016; @thomas_multi-level_2021].

We will go through a simple `threemc` model of an example dataset for Malawi. 
It's important to note that **we only look at a subset of Malawi's surveys** 
here (not all are publicly available), so 
**our results are not reliable and shouldn't be taken as genuine predictions**.

First, load the prerequisite packages, which won't be too many beyond 
`threemc` itself.
```{r setup}
# for analysis
library(dplyr, warn.conflicts = FALSE)
library(sf)
library(threemc)

# for plotting
library(ggplot2)
library(scales)

# Revert to using planar rather than spherical geometry in `sf`
sf_use_s2(FALSE)
```


## Metadata

Next, set the following metadata in our analysis:
```{r metadata}
cntry <- "MWI"          # iso3 code for country to model 
cens_age <- 59          # max circ. age to include in model, censor older ages
N <- 1000               # N samples from posterior predictive distribution
start_year <- 2008      # year to begin predictions 
forecast_year <- 2021   # predictions up to this year

# PSNU area level for modelled country
area_lev <- threemc::datapack_psnu_area_level %>% 
  filter(iso3 == cntry) %>% 
  pull(psnu_area_level)
area_lev
```

For this example, we will use the default  "standard" `threemc` model, 
which does not include a time effect for TMC (i.e. it assumes that TMC 
practises don't and won't change over time), uses an AR 1 temporal process, and 
treats MMC for all ages as time varying. Models with these features will be 
introduced in a later vignette.

Note that `threemc` contains the dataset `datapack_psnu_area_level`, which 
contains the PSNU ("Primary Sampling Unit") area level for surveys conducted in 
many Sub-saharan African countries. For Malawi, this is area level five, which 
corresponds to "Health District + Cities". Each area level has a different 
meaning for each county, but in general, higher values correspond to more 
granular regions.


## Data

Next, we load our required datasets. Note that this data is normally 
loaded using `read_circ_data`, which avoids having to switch between 
`readr::read_csv` and `sf::read_sf`, and makes repeated filtering more terse. 
However, the demo data we will use in this analysis is directly included in 
`threemc`, so we just explicitly assign it to the global environment from 
there.

```{r load_data}
#' # "Normal" data loading procedure (same for areas, surveys and pops)
#' filters = c("iso3" = cntry, "sex" = male)
#' survey_circumcision = read_circ_data(
#'   path    = "path/to/survey_circumcision", 
#'   filters = filters
#' ) 
survey_circumcision <- threemc::demo_survey_circumcision
areas <- threemc::demo_areas %>% 
  mutate(space = row_number()) # Add unique identifier for each area
populations <- threemc::demo_populations
```

For more information on these datasets and what they contain, please see their 
respective help files (using e.g. `help("demo_survey_circumcision)"`). If you 
wish to fit a model using `threemc` with your own data, it must include the 
same columns as these datasets.


## Prepare Survey Data 

We now preprocess our survey data.
```{r preprocess_surveys}
# pull latest censoring year from survey_id
survey_years <- unique(survey_circumcision$survey_id)
cens_year <- max(survey_years)

# Prepare circ data, and normalise survey weights and apply Kish coefficients.
survey_circ_preprocess <- prepare_survey_data(
  areas               = areas,
  survey_circumcision = survey_circumcision,
  area_lev            = area_lev,
  start_year          = start_year,
  cens_year           = cens_year,
  cens_age            = cens_age
)
head(survey_circ_preprocess)
```

Amongst other things, `prepare_survey_data`:   
- Removes rows with missing essential information (e.g. circumcision status),
and identifies surveys which are completely excluded due to missing data,  
- Censors circumcisions at `cens_age` and `cens_year`, for circumcision age 
and year, respectively,  
- sets the desired area level of aggregation for each survey, reassigning 
surveys at a more granular level to the specified `area_lev` (e.g. MWI_6_309 
(STA Tombondiya), a "Traditional Authority" is reassigned to it's 
"Health District + Cities", MWI_5_22 (Mulanje)),  
- Identifies left (i.e. those circumcised at an unknown age, labelled with the 
column `event` == 1) and right (i.e. still uncircumcised, labelled with 
`event` == 2) censored individuals (those uncensored have `event` == 0),  
- Uses `circ_who` and `circ_where` to determine circumcision `type`,  
- Identifies surveys for which this type information is unavailable, and
- normalises survey weights and applies the Kish coefficient.  

In these Malawi surveys, we do see the presence of some left censoring, but 
thankfully every survey contains enough valid circumcision data to be used.


## Create Shell Dataset 

We wish to make predictions for each unique combination of:  
- The `area_id`s in `survey_circ_preprocess`,  
- circumcision age 0-`max(survey_circ_preprocess$circ_age)`,  
- circumcision years from `start_year`-`forecast_year`, and  
- each type of circumcision (TMC, MMC and for "Missing" type, which can 
still be used when modelling Total MC).

To that end, we use `create_shell_dataset` to use our shapefiles and 
preprocessed surveys to create, well, a shell dataset!
```{r create_shell_data}
out <- create_shell_dataset(
    survey_circumcision = survey_circ_preprocess,
    populations         = populations,
    areas               = areas,
    area_lev            = area_lev,
    start_year          = start_year,
    end_year            = forecast_year,
    time1               = "time1",
    time2               = "time2",
    strat               = "space",
    age                 = "age",
    circ                = "indweight_st"
)
head(out)
```

For each of these unique combinations, `create_shell_dataset` also 
adds their respective `population` from `populations`, and calculates:   
- `N`, the person years, 
- `obs_mmc`, `obs_tmc` `obs_mc` (representing those with "Missing" 
circumcision type), `cens` and `icens`, the observed rate of circumcision for 
each circumcision type, 
- Also the rate of right-censoring (uncircumcised) and left-censoring 
(unknown circumcision age).


## Model Fit Matrices

The final step before fitting our model is to prepare our model data. 
The function `threemc_prepare_model_data` outputs a list, and calculates:
- design matrices for fixed effects and temporal, age, space and
  interaction random effects, (the list elements with names 
  beginning with `X_` below),
- integration matrices for selecting the instantaneous hazard rate (The 
  `IntMat`s),  
- survival matrices for MMC, TMC, censored and left censored, and  (the 
  matrices labelled "A" (for the various types of circumcision), `B` (for 
  right-censoring) and `C` (for left-censoring)).
- the precision/adjacency matrix for the spatial random effect (`Q_space`).

```{r prepare_model_data}
# prepare_dataset for modelling
dat_tmb <- threemc_prepare_model_data(
    out      = out,
    areas    = areas,
    area_lev = area_lev
)

names(dat_tmb)
```

The message returned by `threemc_prepare_model_data` informs us that we have 
implicitly chosen an AR 1 temporal prior here because we have not specified 
the `rw_order` parameter.


## Fit Model 

Now we can finally fit our model.  

First, we initialise hyperparameter values using `threemc_initial_pars`. This 
uses the matrices in `dat_tmb` to create a named list of the initial 
hyperparameters needed for our `threemc` model. The defaults can be overridden 
by specifying the `custom_init` parameter with a named list of it's own. 

Next, we run `threemc_fit_model`, which uses `TMB::MakeADFun` to create an 
objective function (the negative log-likelihood) which can then be optimised by 
`nlminb` using BFGS. Using Monte Carlo sampling, we also draw `N` samples from 
the joint posterior distribution, conditional on the optimised 
hyperparameters [@eaton_naomi_2021]. 
```{r fit_model, results= "hide"}
# Initialise `threemc` hyperparameters
parameters <- threemc_initial_pars(dat_tmb)

# fit model with TMB and stats::nlminb optimiser
fit <- threemc_fit_model(
    dat_tmb    = dat_tmb,
    parameters = parameters,
    randoms    = c(
        "u_time_mmc", "u_age_mmc", "u_age_mmc_paed", "u_space_mmc",
        "u_agetime_mmc", "u_agespace_mmc", "u_agespace_mmc_paed",
        "u_spacetime_mmc",
        "u_time_tmc", "u_age_tmc", "u_space_tmc", "u_agespace_tmc"
    ),
    N          = N
)
```

<!-- fit <- readRDS(file = "MWI_TMBObjects_DistrictAgeTime_ByType.rds") -->

Be aware that this can take quite a long time. Just think of all 
the random effects, and interactions between these! You can model using lower 
`area_lev` for a faster fit time in a pinch.

```{r}
names(fit)
```

`fit` contains all the expected outputs from supplying `TMB::MakeADFun` 
objective function to `stats::nlminb`. Among other things, we can have a 
look at the optimised hyperparameters for our model:

```{r}
fit$par
```

Here `logsigma` and `logitrho` refer to variance and covariance 
hyperparameters, respectively.

We can also access `N` samples from the posterior predictive distribution for 
the following quantities:
```{r}
names(fit$sample)
```

- `"rate_mmc", "rate_tmc", "rate", "surv"` refer to the circumcision rates,  
- `"inc_mmc", "inc_tmc", "inc"`, refer to the (non-cumulative) circumcision 
  incidence, and   
- `"cum_inc_mmc", "cum_inc_tmc", "cum_inc"` refer to the cumulative incidence,
  i.e. the circumcision coverage.   
  
These sample matrices have `N` columns, and their rows match up with the rows 
in `out`, **for the modelled `area_lev`** (so not at lower area levels). We 
can use the function `compute_quantiles` to compute lower, middle and upper 
quantiles (default `c(0.025, 0.5, 0.975)`, but can be changed by specifying the
`probs` argument) for the samples for each row in `out` (that has 
`area_level == area_lev` (five)).
```{r quantiles}
# subset to specific area level and calculate quantiles for rates and hazard
out_spec <- compute_quantiles(out, fit, area_lev = area_lev)

out_spec %>% 
  select(-c(N:icens)) %>%  # remove observed circ rate cols we already know about
  head()
```

<!-- out_spec <- readr::read_csv("./MWI_Results_DistrictAgeTime_ByType.csv.gz") -->

The lower, middle and upper quartiles are labelled with "L", "M" and 
"U", respectively.

You can now save `fit` as an `.RData` file using `save`. However, this file is 
often very large (> 1GB). We can make the `fit` object much smaller using 
`minimise_fit_obj`, which just stores your `dat_tmb` and `parameters` in `fit` 
(under the names `tmb_data` and `par_init`, respectively) and removes 
`fit$sample` and `fit$obj`. This object is much smaller, and can be saved using 
`saveRDS` as an `.rds` file. The downside is that if we want to reload our 
`fit_min` at a later time, we must re-sample. However, this can be easily done 
using `threemc_fit_model` and supplying `fit_min` as the `fit` argument. This 
is treated with greater detail in the aggregation vignette.
```{r minimise_fit}
# minimise fit object for saving
fit_min <- minimise_fit_obj(fit, dat_tmb, parameters)
```


## Interpretation

We can have a quick look at our model predictions by plotting some of our 
results. For the sake of brevity, we won't go (too much) into 
differences in circumcision type here, and only look at total circumcision 
(MC). We'll also have to cheat a little bit and look at only a subset of 
our modelled region, in order for our plots to be readable when exported to 
HTML.

First, we look at circumcision rate for each "Health District + Cities" 
modelled for different (circumcision) ages. We also include several years to 
observe any changes over time.
```{r plot_rates, out.width = "100%", out.height = "100%", dpi = 500}
# remove some years to show change more clearly
out_plot <- out_spec %>% 
  filter(
    year %in% c(
      start_year, ceiling((start_year + forecast_year) / 2), forecast_year
    ),
    area_name %in% c(
      "Blantyre", "Blantyre City", "Lilongwe", 
      "Lilongwe City", "Machinga", "Mangochi", "Mzimba North", 
      "Mzimba South", "Zomba City" #  "Chiradzulu", "Mzuzu City", "Zomba"
    )
  )

ggplot(
  out_plot,
  aes(
    x = age,
    y = rateM,
    ymin = rateL,
    ymax = rateU,
    group = as.factor(year),
    colour = as.factor(year)
  )
) +
  geom_ribbon(fill = "lightgrey", colour = NA) +
  geom_line(size = 1) +
  # facet_wrap(. ~ area_name, ncol = 4) + 
  facet_wrap(. ~ area_name) + 
  labs(x = "Age", y = "Rates", colour = "") +
  scale_y_continuous(labels = label_percent()) +
  theme_bw() +
  theme(
    axis.text.x      = element_text(size = 5),
    axis.title.x     = element_text(size = 8),
    axis.text.y      = element_text(size = 5),
    axis.title.y     = element_text(size = 8),
    legend.text      = element_text(size = 5),
    legend.position  = "bottom",  # place legend on bottom
    strip.background = element_rect(fill = NA, colour = "white") # col facets
  )
```

There is significant spatial heterogeneity in circumcision practises. However, 
we can see that there is not much heterogeneity when it comes to circumcision 
age, with the MC rate highest among those aged around 10-13 for every area with 
less than negligible circumcision. Circumcisions do not seem to occur for those 
over 30s.

In many areas (such as Lilongwe), there doesn't seem to be a great number of 
circumcisions being performed. Perhaps this is because circumcision is already 
high in these regions. However, this may be unlikely, as traditional 
circumcision (the main reason for high circumcision in Malawi pre-2012 (the 
year they began implementing Voluntary Male Medical Circumcision (VMMC) 
programmes [@matoga_uptake_2022])) is generally low in these regions (this can 
easily be plotted using `rate_tmc`). 

Also, circumcision rates have increased quite 
significantly in many regions since 2008. However, in some (such as Blantyre 
City and Machinga), MC rates have slowed in 2021 (for which we have no survey 
data and must forecast out-of-sample) vs 2015, indicating that circumcision 
coverage may have become quite high and started to "taper-off" there.

It's also obviously a good idea to look at circumcision coverage to help 
support or reject our theories, and give additional insight.
```{r plot_coverage, out.width = "100%", out.height = "100%", dpi = 500}
ggplot(
  out_plot,
  aes(
    x = age,
    y = cum_incM,
    ymin = cum_incL,
    ymax = cum_incU,
    group = as.factor(year),
    colour = as.factor(year)
  )
) +
  geom_ribbon(fill = "lightgrey", colour = NA) + 
  geom_line(size = 1) +
  facet_wrap(. ~ area_name) + 
  labs(x = "Age", y = "Coverage", colour = "") +
  scale_y_continuous(labels = label_percent()) + 
  theme_bw() + 
  theme(
    axis.text.x      = element_text(size = 5),
    axis.title.x     = element_text(size = 8),
    axis.text.y      = element_text(size = 5),
    axis.title.y     = element_text(size = 8),
    legend.text      = element_text(size = 5),
    legend.position  = "bottom",  
    strip.background = element_rect(fill = NA, colour = "white") 
  )
```

Here we can see the high MC rates amongst younger individuals reflected in 
the age-distribution of circumcision coverage in many of these same regions, 
which seems to "peak" at slightly older teenage years (around 17-19).

In Lilongwe and other regions with "flat" MC rates in our previous plot, we 
unfortunately learn that circumcision coverage is very low.

Circumcision coverage is greater than or the same in every region from 
2008 to 2015 and from 2015 to 2021 (however, this is one of the assumptions 
of `threemc`, that MC is cumulative thus monotonically increasing).

In those regions where we observed some "tapering off" in circumcision rates, 
we can see that coverage is indeed high. In particular, Machinga and Mangochi 
have had consistently high coverage across our predictive period, suggesting 
that TMC is high there. Blantyre City and Lilongwe City, which have high HIV 
burdens [@nutor_spatial_2020] and had very low coverage in 2008, have seen 
significant (forecasted) increases in circumcision levels, highlighting the 
work being done in resource allocation, which motivates this model 
(this perhaps also suggests a greater ease in implementing circumcision 
programmes in urban areas in Malawi).


## Conclusions

We have fitted a `threemc` model to Malawi's "Health District and Cities", 
and interpreted our results (which, again, are 
**not to be taken as genuine forecasts for Malawi**). Hopefully this vignette 
can empower you to fit `threemc` to your own data!

We can also aggregate these results to less granular regions, producing 
estimates for each of Malawi's five area levels, including at the 
national level. This is the focus of the next vignette.

There are also several variations on the "default" `threemc` model used here. 
As mentioned previously, these include having a fixed paediatric MMC rate, a 
time effect for TMC, and using RW temporal priors. These alternative models are 
also addressed in another vignette.

## References 

<div id="refs"></div>